### 2021-07-09 rad-icfp-2021/intro.tex
Typos and fixes.

\begin{example}[Simple operations on arrays]
    On arrays, three simple operations of interest are the dot-product of two vectors, and the product or sum of the elements of a vector.
    In a functional setting, these can be defined as follows:
\begin{center}
    \begin{tabular}{{r c l}}
        !prod(A)! &:=& !reduce * 1 A! \\
        !sum(A)! &:=& !reduce + 0 A! \\
        !dot(A,B)! &:=& !reduce + 0 (map2 * A B)!     
    \end{tabular}
\end{center}
where !reduce! is a known fold-left operator for which the function argument is associative. 
It is notably faster to execute than a fold-left as it is parallel friendly.

The gradient of each of these expressions w.r.t. !A! will be:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))! \\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where:
\begin{itemize}
\item !scanl! is the scan-left operator that returns all the intermediate results of fold-left,
\item !scanr! is the scan-right operator that returns all the intermediate results of fold-right,
\item !shift1L [v$_1$,$\ldots$,v$_n$]! is the shift-left operator and returns ![v$_2$,$\ldots$,v$_{n}$]!, and 
\item !shift1R [v$_1$,$\ldots$,v$_n$]! is the shift-right operator and returns ![v$_1$,$\ldots$,v$_{n-1}$]!.
\end{itemize}
These gradients are some examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

@@

\begin{itemize}[leftmargin=*]
\item We present our work with a simple yet expressive array-based language (with constructs such as !map2! and !reduce!) in Section~\ref{sec:simplediff}. 
We show how to directly compute an efficient reverse-mode AD for the expressions of this program (top of Figure~\ref{fig:outline}).  
Furthermore, we show how to extend our work to a richer language in Section~\ref{sec:generalization}.
\item One of the key insights behind efficient reverse-mode AD is to only consider unary operators. 
Inspired by this insight and following IRs such as SSA and ANF, we introduce a novel IR, UNF (Section~\ref{sec:unf})).
We introduce an alternative and easier to follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
\item We prove the complexity guarantees for the reverse ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
\item We proof the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining the denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
\end{itemize}

>>>>

\begin{example}[Simple operations on arrays]
    On arrays, three simple operations of interest are the dot-product of two vectors, and the product or sum of the elements of a vector.
    In a functional setting, these can be defined as follows:
\begin{center}
    \begin{tabular}{{r c l}}
        !prod(A)! &:=& !reduce * 1 A! \\
        !sum(A)! &:=& !reduce + 0 A! \\
        !dot(A,B)! &:=& !reduce + 0 (map2 * A B)!     
    \end{tabular}
\end{center}
where !reduce! is a known fold-left operator for which the function argument is associative. 
It is notably faster to execute than a fold-left as it is parallel-friendly.

The gradient of each of these expressions with respect to !A! is:
\begin{center}
    \begin{tabular}{{r c l}}
        $\nabla_A$!prod(A)! &:=& !map2 * (scanr * 1 (shift1L A)) (shift1R (scanl * 1 A))! \\
        $\nabla_A$!sum(A)! &:=& !map (x -> 1) A!\\
        $\nabla_A$!dot(A,B)! &:=& !B! 
    \end{tabular}
\end{center}
where
\begin{itemize}
\item !scanl! is the scan-left operator that returns all the intermediate results of fold-left,
\item !scanr! is the scan-right operator that returns all the intermediate results of fold-right,
\item !shift1L [v$_1$,$\ldots$,v$_n$]! is the shift-left operator and returns ![v$_2$,$\ldots$,v$_{n}$]!, and 
\item !shift1R [v$_1$,$\ldots$,v$_n$]! is the shift-right operator and returns ![v$_1$,$\ldots$,v$_{n-1}$]!.
\end{itemize}
These gradients are a few examples among numerous ones which are usually derived by hand, and are here obtained automatically as special cases of our work.
\end{example}

@@

\begin{itemize}[leftmargin=*]
\item We present our work with a simple yet expressive array-based language (with constructs such as !map2! and !reduce!) in Section~\ref{sec:simplediff}. 
We show how to directly compute an efficient reverse-mode AD for the expressions of this program (top of Figure~\ref{fig:outline}).  
Furthermore, we show how to extend our work to a richer language in Section~\ref{sec:generalization}.
\item One of the key insights behind efficient reverse-mode AD is to only consider unary operators. 
Inspired by this insight and following IRs such as SSA and ANF, we introduce a novel IR, UNF (Section~\ref{sec:unf})).
We introduce an alternative and easier-to-follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
\item We prove complexity guarantees for the reverse-ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
\item We prove the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining a denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
\end{itemize}
